---
title: "Getting Started with RII"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RII}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What RII is, and why this package exists

The Relative Importance Index (RII) ranks a set of factors, causes or
criteria that respondents have rated on a Likert scale. For item $j$:

$$RII_j = \frac{\sum W_j}{A \times N_j}$$

where $\sum W_j$ is the sum of the weights (raw Likert scores) given to
item $j$ by all respondents, $A$ is the highest possible weight (5 on a
5-point scale), and $N_j$ is the number of respondents who rated it. RII
always falls between 0 and 1, which makes items directly comparable and
rankable — this is why it is so widely used in environmental,
construction and social-science surveys.

Most researchers compute this in a spreadsheet: one `SUMPRODUCT` and one
`COUNT` formula per item, copied down a column, then sorted by hand. The
**RII** package does the same calculation from R, with automatic ranking
and a ready-made chart, as a first step towards analysing survey data in
R instead of Excel.

## The example data

The package ships a worked Nigerian example: 150 construction
professionals across seven states rating ten perceived causes of building
collapse on a 5-point scale (1 = Not significant, 5 = Extremely
significant).

```{r setup}
library(RII)
data(building_collapse_nigeria)
head(building_collapse_nigeria[, 1:6])
```

## Ranking items from raw responses

If you have one row per respondent and one column per item — the usual
shape for data exported from Google Forms, KoboToolbox or a survey
platform — pass those columns straight to `rii_table()`:

```{r}
items <- building_collapse_nigeria[, 5:14]
tbl <- rii_table(items)
tbl
```

The result is sorted by RII automatically, and each item is placed into
one of the five importance bands used in the literature (Low,
Medium-Low, Medium, High-Medium, High) via `classify_rii()` internally.

`summary()` gives a quick narrative reading of the table:

```{r}
summary(tbl)
```

## A one-line entry point for messy real-world survey data

Real exports usually mix Likert items with demographic/ID columns
(gender, age, marital status...) that are numerically just as
"Likert-like" as a real item — a gender column coded 1/2 looks no
different to R than a real 2-point item. Typing out 30+ item column
names by hand is tedious, so `items = "auto"` detects them from the
common short-code naming convention (`OJ1`, `PJ6`, `JS1i`, ...) instead:

```{r}
set.seed(1)
raw_survey <- data.frame(
  `S/N` = 1:40, GENDER = sample(1:2, 40, TRUE),
  MARITAL = sample(1:3, 40, TRUE),
  OJ1 = sample(1:5, 40, TRUE), OJ2 = sample(1:5, 40, TRUE),
  PJ6 = sample(1:5, 40, TRUE), JS1i = sample(1:5, 40, TRUE),
  check.names = FALSE
)

# review the guess first
detect_likert_items(raw_survey)

# or just let rii_table() use it directly
rii_table(raw_survey, items = "auto")
```

## Multiple constructs/sections in one questionnaire

A longer questionnaire often has several distinct sections — e.g. "Road
Surface Quality" (columns 4-10) and "Maintenance Practices" (columns
11-13) — that should usually be ranked **separately**, since mixing two
different constructs into one RII table isn't generally meaningful.
`items` accepts a numeric column range for exactly this, so you can call
`rii_table()` (or `read_rii_csv()`/`read_rii_excel()`/`read_rii_spss()`)
once per section:

```{r, eval = FALSE}
d <- read.csv("road_infra_survey_raw.csv")   # read the whole file once

road_quality <- rii_table(d, items = 4:10)   # Section 1
maintenance  <- rii_table(d, items = 11:13)  # Section 2
```

Note that the column range has to be a real R argument (`items = 11:17`),
not text appended to the file path — `read_rii_csv("file.csv[, 11:17]",
...)` won't work, since indexing isn't part of a file name.

## Text Likert labels ("Strongly Agree", "Disagree", ...)

Many surveys (Google Forms exports especially) record responses as text
rather than numbers. `recode_likert()` converts them, and can auto-detect
which columns to convert:

```{r}
text_survey <- data.frame(
  Gender = sample(c("Male", "Female"), 30, TRUE),
  Q1 = sample(c("Strongly disagree", "Disagree", "Neutral", "Agree", "Strongly agree"), 30, TRUE),
  Q2 = sample(c("Strongly disagree", "Disagree", "Neutral", "Agree", "Strongly agree"), 30, TRUE)
)

# review which columns it would recode, and using which scale
detect_text_likert_items(text_survey)

# recode, then rank — items = "auto" picks up exactly the recoded columns
text_survey2 <- recode_likert(text_survey)
rii_table(text_survey2, items = "auto")
```

Built-in wordings cover agreement (with or without "Neutral"/"Undecided"),
frequency, satisfaction, importance and extent scales. For anything else,
pass your own mapping: `recode_likert(d, cols = "Q1", mapping = c("no" =
1, "maybe" = 2, "yes" = 3))`.

## Ranking items from a frequency table

Many spreadsheet RII templates don't keep raw respondent-level data —
only a count of how many respondents chose each scale point for each
item. `rii_table_freq()` works directly from that layout:

```{r}
freq <- data.frame(
  Factor = c("Poor supervision", "Use of substandard materials"),
  SD = c(2, 1), D = c(5, 4), N = c(10, 12), A = c(38, 40), SA = c(95, 93)
)
rii_table_freq(freq, item_col = "Factor")
```

## Reading straight from an Excel workbook

If your data already lives in an `.xlsx` file — either layout — you can
skip the import step entirely:

```{r}
raw_path <- system.file("extdata", "rii_raw_example.xlsx", package = "RII")
read_rii_excel(raw_path, layout = "raw",
                items = c("SubstandardMaterials", "PoorSupervision",
                          "InadequateSoilInvestigation", "PoorWorkmanship"))
```

## Writing results back to Excel

If you'd rather hand someone a spreadsheet than an R object, write the
ranked table straight to `.xlsx`. Importance bands are colour-coded the
way many published RII tables shade them, and the chart is embedded in
the same sheet by default:

```{r, eval = FALSE}
out <- tempfile(fileext = ".xlsx")
write_rii_excel(tbl, out)
```

## Plotting the ranking

```{r, fig.alt = "Horizontal bar chart of RII values, sorted from highest to lowest"}
plot(tbl, main = "Causes of Building Collapse in Nigeria")
```

Use `top_n` to show only the highest-ranked items, and `col` to change
the bar colour for a different context:

```{r, fig.alt = "Horizontal bar chart of the top five RII values"}
plot(tbl, top_n = 5, col = "steelblue",
     main = "Top 5 Causes of Building Collapse")
```

## Single-item building blocks

If you want to build something custom, `rii_raw()` and `rii_freq()`
compute the same statistics for one item at a time:

```{r}
rii_raw(items$PoorSupervision)
rii_freq(c(2, 5, 10, 38, 95))
```

## A non-construction example, on a different scale

RII is not specific to construction, and neither is this package. The
second bundled dataset, `csa_adoption_barriers_nigeria`, is a survey of
smallholder farmers rating barriers to climate-smart agriculture adoption
— on a **4-point** scale rather than 5-point, to show that `scale_max`
(and `scale_min`, for a 0-based or otherwise shifted scale) adapts to
whatever instrument was actually used:

```{r}
data(csa_adoption_barriers_nigeria)
csa_items <- csa_adoption_barriers_nigeria[, 6:15]
csa_tbl <- rii_table(csa_items, scale_max = 4)
csa_tbl
```

```{r, fig.alt = "Horizontal bar chart of RII values for barriers to climate-smart agriculture adoption"}
plot(csa_tbl, main = "Barriers to Climate-Smart Agriculture Adoption",
     col = "#6B8E23")
```
